### 2015-11-17 paper.Rmd
objects!

It is clear that census-type surveys (quadrat or strip transects) are inefficient (requiring a lot of effort on the part of those in the field) and we should expect that not all animals can be observed. Accounting for imperfect detectability is an important consideration when want to obtain accurate estimates of abundance [@LahozMonfort:2013bg]. Using the extra information gained by recording distance from the sampler to the observation, it is possible to model detectability. Since we expect detectability to decrease with increasing distance from the sampler, we model detectability as a function of distance (plus perhaps other covariates, see below).

@@

## line transects
opar2 <- par(mar=c(1, 1, 1, 1) + 0.1)
plot(x, y, pch=19, asp=1, cex=0.6, main="",col="grey", axes=FALSE, xlab="", ylab="")
polygon(x=c(0,0,1,1,0), y=c(0,1,1,0,0))
# generate some lines
# in this case we don't randomise the offset of the grid
lt <- list(x=c(-0.0125,-0.0125,0.0125,0.0125,-0.0125), y=c(0,1,1,0,0))
# set sigma
sigma <- 0.01
# storage for detected distances
detected_distances <- c()
for(i in 1:5){
  # calculate next strip location
  lt$x <- lt$x+0.15
  # plot the line transect
  lines(x=rep(mean(range(lt$x)),2), y=c(0,1), col="blue",lty=2)
  # calculate the distances to animals from the line
  distances <- abs(lt$x - x)
  # randomly decide which were detected
  detected <- exp(-distances^2/(2*sigma^2)) > runif(length(distances))
  # plot those animals detected
  points(x[detected], y[detected], pch=19, cex=0.6, col="red")
  # collect the distances to detected objects
  detected_distances <- c(detected_distances, distances[detected])
}

@@

# set sigma
sigma <- 0.05
# storage for detected distances
detected_distances <- c()
# lay out a grid of points
pt <- as.list(expand.grid(x=seq(0.15, 0.85, len=3), y=seq(0.15, 0.85, len=3)))
for(i in 1:length(pt$x)){
  # generate point location
  # plot the line transect
  points(pt$x[i], pt$y[i], pch=19, col="blue",cex=0.7)
  # calculate the distances to animals from the line
  distances <- sqrt((pt$x[i] - x)^2+(pt$y[i]-y)^2)
  # randomly decide which were detected
  detected <- exp(-distances^2/(2*sigma^2)) > runif(length(distances))
  # plot those animals detected
  points(x[detected], y[detected], pch=19, cex=0.6, col="red")
  # collect the distances to detected objects
  detected_distances <- c(detected_distances, distances[detected])
}

@@

The rest of the paper is structured as follows: we first look into how data needs to be organised to allow it to be used with \pkg{Distance}; models for the detection function are then covered including their formulation and examples of fitting in \proglang{R}. We then spend some time looking at model checking and goodness of fit testing, as well as model selection. Having illustrated how to obtain a good detection function, we show how to estimate abundance using that model, including using post-stratification. The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for animal abundance estimation.

@@

* *Shoulder*: we expect observers to be able to see objects near them, not just those directly infront of them. For this reason, we expect the detection function to be "flat" near zero distance.
* *Non-increasing*: we don't think that observers should be more likely to see things further away than those nearer to them. This usually indicates an issue with field procedure (that the distribution of animals with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
* *Model robust*: models should be flexible enough to make many different shapes.
* *Pooling robust*: many factors can affect the probability of detection and it is not possible to measure all of these. We would like our models to be robust to us not including these factors.
* *Estimator efficiency*: we would like our models to have low variances, but only given they satisfy the other properties above (which, if they are satisfied, would give low bias).

@@

To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot). In our case the CDF evaluates the probability of observing an animal at a distance less than or equal to some given value. The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance. In essence we're asking "is the number of observations up to a given distance in line with what the model says they should be?" (where “given values” are the observed distances). As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.

@@

1. *Model uncertainty*, from the estimation of the detection function parameters $\boldsymbol{\theta}$
2. *Sampling uncertainty*, from the distribution of animals along the transect lines or between visiting occasions for points.

@@

Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: the number of animals per unit transect. When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits to all points). When covariates are included in the detection function, it is recommended that we substitute the $n$ in the previous expressions with the estimated abundance $\hat{N}$ as this will take into account the effects of the covariates.

>>>>

It is clear that census-type surveys (quadrat or strip transects) are inefficient (requiring a lot of effort on the part of those in the field) and we should expect that not all objects (animals, plants, dung, etc) can be observed. Accounting for imperfect detectability is an important consideration when want to obtain accurate estimates of abundance [@LahozMonfort:2013bg]. Using the extra information gained by recording distance from the sampler to the observation, it is possible to model detectability. Since we expect detectability to decrease with increasing distance from the sampler, we model detectability as a function of distance (plus perhaps other covariates, see below).

@@

## line transects
opar2 <- par(mar=c(1, 1, 1, 1) + 0.1)
plot(x, y, pch=19, asp=1, cex=0.6, main="",col="grey", axes=FALSE, xlab="", ylab="")
polygon(x=c(0,0,1,1,0), y=c(0,1,1,0,0))
# generate some lines
# in this case we don't randomise the offset of the grid
lt <- list(x=c(-0.0125,-0.0125,0.0125,0.0125,-0.0125), y=c(0,1,1,0,0))
# set sigma
sigma <- 0.01
# storage for detected distances
detected_distances <- c()
for(i in 1:5){
  # calculate next strip location
  lt$x <- lt$x+0.15
  # plot the line transect
  lines(x=rep(mean(range(lt$x)),2), y=c(0,1), col="blue",lty=2)
  # calculate the distances to objects from the line
  distances <- abs(lt$x - x)
  # randomly decide which were detected
  detected <- exp(-distances^2/(2*sigma^2)) > runif(length(distances))
  # plot those objects detected
  points(x[detected], y[detected], pch=19, cex=0.6, col="red")
  # collect the distances to detected objects
  detected_distances <- c(detected_distances, distances[detected])
}

@@

# set sigma
sigma <- 0.05
# storage for detected distances
detected_distances <- c()
# lay out a grid of points
pt <- as.list(expand.grid(x=seq(0.15, 0.85, len=3), y=seq(0.15, 0.85, len=3)))
for(i in 1:length(pt$x)){
  # generate point location
  # plot the line transect
  points(pt$x[i], pt$y[i], pch=19, col="blue",cex=0.7)
  # calculate the distances to objects from the line
  distances <- sqrt((pt$x[i] - x)^2+(pt$y[i]-y)^2)
  # randomly decide which were detected
  detected <- exp(-distances^2/(2*sigma^2)) > runif(length(distances))
  # plot those objects detected
  points(x[detected], y[detected], pch=19, cex=0.6, col="red")
  # collect the distances to detected objects
  detected_distances <- c(detected_distances, distances[detected])
}

@@

The rest of the paper is structured as follows: we first look into how data needs to be organised to allow it to be used with \pkg{Distance}; models for the detection function are then covered including their formulation and examples of fitting in \proglang{R}. We then spend some time looking at model checking and goodness of fit testing, as well as model selection. Having illustrated how to obtain a good detection function, we show how to estimate abundance using that model, including using post-stratification. The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for estimating the abundance of biological populations.

@@

* *Shoulder*: we expect observers to be able to see objects near them, not just those directly infront of them. For this reason, we expect the detection function to be "flat" near zero distance.
* *Non-increasing*: we don't think that observers should be more likely to see things further away than those nearer to them. This usually indicates an issue with field procedure (that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
* *Model robust*: models should be flexible enough to make many different shapes.
* *Pooling robust*: many factors can affect the probability of detection and it is not possible to measure all of these. We would like our models to be robust to us not including these factors.
* *Estimator efficiency*: we would like our models to have low variances, but only given they satisfy the other properties above (which, if they are satisfied, would give low bias).

@@

To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot). In our case the CDF evaluates the probability of observing an object at a distance less than or equal to some given value. The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance. In essence we're asking "is the number of observations up to a given distance in line with what the model says they should be?" (where “given values” are the observed distances). As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.

@@

1. *Model uncertainty*, from the estimation of the detection function parameters $\boldsymbol{\theta}$
2. *Sampling uncertainty*, from the distribution of objects along the transect lines or between visiting occasions for points.

@@

Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: the number of objects per unit transect. When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits to all points). When covariates are included in the detection function, it is recommended that we substitute the $n$ in the previous expressions with the estimated abundance $\hat{N}$ as this will take into account the effects of the covariates.
